---
title: "REVISED_SDS300CR_Lab_2_Danielle_Weiss"
author: "Danielle Weiss"
format:
  html:
    embed-resources: true
    toc: true
  pdf:
    toc: true
    number-sections: true
    colorlinks: true
    self-contained: true
editor: source
---


# **Part 0:** Read in packages we need

Make a code chunk below and read in any packages you may need.

```{r}
library(dplyr)
library(tidyverse) 
library(ggplot2) # plots
library(ggsci) # colors
library(palmerpenguins) # penguins data
library(performance) # stat + model assumptions [check_model]
library(rstatix) #test for outliers, welch_anova_test
library(lubridate) # data and time
``` 


# **Part 1:** Reshape data and test the Intermediate Disturbance Hypothesis

1. Read in [this data](https://github.com/jbaumann3/BIOL234_Biostats_MHC/blob/main/Ocean_Point_Data%202018.csv) from github
```{r}
# read in github data csv - RAW DATA
ocean_point_data <- read_csv("https://raw.githubusercontent.com/jbaumann3/BIOL234_Biostats_MHC/main/Ocean_Point_Data%202018.csv")

# check loaded data
head(ocean_point_data)
```

2. Today we are interested in how much algal diversity is in a place given the number of grazers present. In short- is there a relationship between grazers (snail) and algae? Bare rock, carcinus, and cancer crabs columns can be dropped as we are not using them. Do this below
```{r}
#drop columns we don't need!
ocean_point <- select(ocean_point_data, -c('bare_rock', 'carcinus', 'Cancer crabs'))

# check new data
head(ocean_point)
```


3. The data are survey data from the rocky intertidal in coastal Maine. Students worked at a transect (trans_description) in different years (year), they measured how high they were above (or below) the low tide line (tidal_ht), those numerical tide heights were converted into a ranked list to make bins (tidal_height_cat_number, 1-8 or so), and then students recorded either the percent of 2D area that was covered by a thing (if it was something that didn't move) OR how many individuals of a thing they saw (if a crab or snail or something). This abundance or cover data is in columnar form by species (or category) in the remaining columns. This works nicely when we are recording data on a clipboard in the field. BUT, R wants all of those names in 1 column and all of those numbers in a second column. We can use pivot_longer to convert this dataframe to long format. 

Pivot the data to long format - one column for names and one for values
```{r}
# pivot wide to long 
ocean_point_long <- (ocean_point %>%
  pivot_longer(cols = calothr:nucella, 
               names_to = 'species', 
               values_to = 'abundance_or_cover'))

# check new data
head(ocean_point_long)
```


4. Make a new column to tell us if value = %cover or abundance. The grazers (animals) are counted in abundance and the algaes (everything else) are % cover. You may also want to make an algae vs grazer column. Do this below: 

removed these:
bare_rock 
Cancer crabs
carnicus

species being used:
tectura abundance
litt_lit abundance
litt_obt abundance
litt_sax abundance
nucella abundance
algae % cover
mytilus % cover - animal 
semibal % cover - animal

```{r}
# make new column
ocean_point_long$measurement_type <- ifelse(ocean_point_long$species == "tectura" | ocean_point_long$species == "litt_lit" | ocean_point_long$species == "litt_obt" | ocean_point_long$species == "litt_sax" | ocean_point_long$species == "nucella", "abundance", "%cover")

# check new data
head(ocean_point_long)
ocean_point_long
```

5. We are interested in a visual that assess the intermediate disturbance hypothesis, which states that there is some middle range of disturbance (grazing in this case) that leads to the most DIVERSE algal ecosystem. Calculate total grazer count and total algal diversity for each tidal height cat num. 
```{r}
# get rid of mytilus + semibal 
ocean_points_long <- ocean_point_long %>% filter(!(tolower(species) %in% c('mytilus', 'semibal')))

# make sure they are removed
ocean_points_long

# total grazer count for each tidal height group - only include abundance data - grazers only 
# abundance_or_cover - numerical values of abundance # or %cover 
grazer_counts <- ocean_points_long %>%
  filter(measurement_type == "abundance") %>%
  group_by(tidal_ht_cat_num) %>%
  summarize(total_grazer_count = sum(abundance_or_cover))

# total algal diversity for each tidal height group - only include % cover data - all algae species - exclude species with 0% cover (not present)
algal_diversity <- ocean_points_long %>%
  filter(measurement_type == "%cover" & abundance_or_cover > 0) %>%
  group_by(tidal_ht_cat_num) %>%
  summarize(total_algal_diversity = n_distinct(species))

# check new data
grazer_counts
algal_diversity

# combine dataframes to include algal_diversity and grazer_counts
gzcount_algdiv <- full_join(grazer_counts, algal_diversity, by = "tidal_ht_cat_num")

# check new data 
gzcount_algdiv
```

6. Graph to visually assess the IDH-- grazer abundance on the X and algal diversity on the y!
What trends do you see? Do you see any evidence for the IDH? Why or why not?
```{r}
# make plot to assess IDH hypothesis
idh_plot_div_abun <- ggplot(gzcount_algdiv, aes(x = total_grazer_count, y = total_algal_diversity, color = factor(tidal_ht_cat_num))) +
  geom_point() +
  geom_line(color = "black") +
  labs(x = "Grazer Abundance", y = "Algal Diversity") +
  ggtitle("Algal Diversity in Response to Grazer Abundance") +
  theme_classic() + 
  scale_y_continuous(breaks = seq(0, max(gzcount_algdiv$total_algal_diversity), 1))

# view plot
idh_plot_div_abun

# Trends in IDH graph 
# there are 4 points at (0,1) with 0 grazers and 1 algal species
# there is no clear pattern or parabola in this relationship of algal diversity vs. grazer abundance
# algal diversity ranged from 1-10 out of the 19 algal species surveyed 
# some tidal heights had no grazers present, and still had low algal diversity - the dominant algal species could have taken over space through competition? 
```


# **Part 2**: dplyr/tidyverse pipelines and ANOVA

1. Read in [this coral cover data](https://github.com/jbaumann3/BIOL234_Biostats_MHC/blob/main/coralcover.csv) from github.

In the following chunk, starting with the raw data from github, build a dplyr/tidyverse pipeline that calculates mean coral cover by reef type. Don't forget to calculate a relevant and useful error term for plotting
```{r}
# read in github data csv - RAW DATA
coral_cov_data <- read_csv('https://github.com/jbaumann3/BIOL234_Biostats_MHC/raw/main/coralcover.csv')

# check data
head(coral_cov_data)

# pipeline for mean coral cover BY reef type 
# mean cc_percent + type 
# use standard error - SE
mean_coralcover_reef <- coral_cov_data %>%
  group_by(type) %>%
  summarize(mean_coralcover = mean(cc_percent), 
            sd = sd(cc_percent), n = n(), se = sd/sqrt(n))

# view data 
mean_coralcover_reef
  
```

2.Make a visual that allows us to assess the following Null Hypothesis: 
There is no difference in coral cover by reef type.
You should make a graph that shows relevant means, error bars, and raw data

```{r}
# raw data points = geom_jitter [coral_cov_data]
# mean coral cover by reef type = geom_point [mean_coralcover_reef]
# SE = geom_errorbar -> measure spread  [standard error of mean coral cover] [mean_coralcover_reef] [raw data points behind means + SE]
# set y limit to 35 to zoom in -> cutting off 1 outlier
coralcover_reeftype <- ggplot()+
  geom_jitter(data = coral_cov_data, aes(x = type, y = cc_percent, color = type), alpha = 0.5, width = 0.2) +
  geom_point(data = mean_coralcover_reef, aes(x = type, y = mean_coralcover, color = type), size = 3) + 
  geom_errorbar(data = mean_coralcover_reef, aes(x = type, ymin = mean_coralcover - se, ymax = mean_coralcover + se), width = 0.1) +
  labs( x = "Reef Type", y = "Mean Coral Cover (%)", color = "Reef Type") + 
  ggtitle("Mean Coral Cover of Reef Types") + 
  theme_bw() +   
  scale_y_continuous(limits = c(0, 35))

# view graph
coralcover_reeftype
# patch reef + black reef error bars overlap 
# no reef overlaps with nearshore 
```

3. Run the correct statistical test to accompany the plot above. This should be a One-Way ANOVA + TukeyHSD test. Run these below AND interpret them. What do you see evidence for? Do we reject or fail to reject the null?

```{r}
# One-Way ANOVA
anova_coral_cov <- aov(cc_percent ~ type, data = coral_cov_data)

# view results
anova_coral_cov
# may have unbalanced effects
summary(anova_coral_cov)
# Pr(>F) = 0.00215 suggests significant differences within means of groups - need TukeyHSD to see which groups are / are not significantly different from one another 

# significance = need post-hoc test (tukey) = what groups are different than other groups
# TukeyHSD test
tukey_coral_cov <- TukeyHSD(anova_coral_cov)

# view results
tukey_coral_cov
# p adj values < 0.05 for Nearshore-Back Reef [0.0072335] and Patch Reef-Nearshore [0.0027941] means these groups have significantly different coral covers from one another [nearshore and black reef are significantly different in coral cover + nearshore and patch reef are significantly different in coral cover] 
# p adj value > 0.05 for Patch Reef-Back Reef [0.9393628] means that patch reef and black reef coral covers are not significantly different 

# Null Hypothesis: There is no difference in coral cover by reef type
# we have evidence [explained above] that we should reject the null hypothesis because there are significant differences in coral cover by reef type in 2/3 of these reef type groups
```
*Revision*
4. Load the penguins data from the palmerpenguins package. Do a two-way ANOVA to assess the effects of species and binary assigned sex at birth (aka the 'sex' column here) on beak length. This should be an interactive two-way ANOVA (Y~X*Z). Make a graph and run the stats. Assess the result using the following H0: "There is no effect of sex or species and there is no interaction between sex and species on penguin beak length"

```{r}
# data 
head(penguins)

# 2 way ANOVA
# y ~ x * z [x and z = independent variables | y = dependent variable]
penguinz <- drop_na(penguins)
penguin_aov1 <- aov(bill_length_mm ~ species * sex, data = penguinz)
summary(penguin_aov1)

# check model assumptions
check_model(penguin_aov1)

# graph: means + error bars (SEM or CI)
# calculate a mean bill length + SEM for penguins by species and sex
beak_lengths_m <- penguinz %>%
  group_by(species, sex) %>%
  summarize(meanbeak = mean(bill_length_mm), sd = sd(bill_length_mm), n = n(), se = sd/sqrt(n)) 

beak_lengths_m

# graph means and SEM 
pen_plot <- ggplot(data = beak_lengths_m, aes(x = species, y = meanbeak, color=sex)) +
    geom_jitter(data = penguinz, aes(x = species, y = bill_length_mm, color = sex), alpha = 0.5, width = 0.2)+ # raw data
  geom_point(data = beak_lengths_m, aes(x = species, y = meanbeak, color = sex), size = 3) + # averages
  geom_errorbar(data = beak_lengths_m, aes(x = species, ymin = meanbeak-se, ymax = meanbeak + se),color = "black", width = 0.1) + 
  labs( x = "Species", y = "Mean Beak Length", color = "Sex") + 
  ggtitle("Mean Beak Lengths of Penguin Species by Sex") +
  theme_classic()

# view plot 
pen_plot
```
*Revision*
4. Stats

```{r}
# summary of interactive 2 way ANOVA
summary(penguin_aov1)

# Null Hypothesis: "There is no effect of sex or species and there is no interaction between sex and species on penguin beak length"
# there is a significant effect of sex on penguin beak length [Pr(>F) = <2e-16 ***]
# there is a significant effect of penguin species on beak length [Pr(>F) = <2e-16 ***]
# there is no significant interaction between sex and species on penguin beak length [Pr(>F) = 0.103]

# TukeyHSD to compare means of beak length based on penguin species or sex
TukeyHSD(penguin_aov1)

# Null Hypothesis: "There is no effect of sex or species and there is no interaction between sex and species on penguin beak length"
# $species
# there is a significant effect of penguin species on beak length for all 3 species interactions
  # Chinstrap-Adelie p-adj = 0.0000000
  # Gentoo-Adelie p-adj = 0.0000000
  # Gentoo-Chinstrap p-adj = 0.0010847
# $sex - male-female 
# there is a significant effect of penguin sex on beak length
  # p-adj = 0.0
# $`species:sex`
# there is 1 instance of no significant interaction between sex and species on penguin beak length
  # Gentoo:female-Chinstrap:female -> p-adj = 0.3338130
  # for female Gentoo and Chinstrap species, there is overlap in the mean error bars     and the p-adj value from the raw data TukeyHSD is 0.33 showing there is no significant difference in beak lengths of female Gentoo and female Chinstrap penguin species
# $`species:sex`
# there is a significant interaction between sex and species on penguin beak length 
  # all other 'species:sex' p-adj values are < 0.05, with no overlapping error bars, so there is a significant difference in beak lengths for these species and sex interactions

```


# **Part 3:** Time series data!

Generally speaking, R has a tough time dealing with time formatting. This is a common issue in programming. There are special data formats called Datetime that we can use to work through this. 

1. Read in [this time series data](https://github.com/jbaumann3/Belize-RT-Baumann-et-al-2021/blob/main/temperature%20and%20light/RT_in_situ_temp.csv) from github
```{r}
# read in github data csv - RAW DATA
temp_light_data <- read_csv("https://github.com/jbaumann3/Belize-RT-Baumann-et-al-2021/raw/main/temperature%20and%20light/RT_in_situ_temp.csv")

# check loaded data
head(temp_light_data)
temp_light_data
```
These data are from an experiment I did recently on coral reefs in Belize. You should see temperature data from 2 sites across a range of dates. 

2. Make a new column the combines the date and time into 1 column, name that column datetime

```{r}
# date time formatting [need to run entire script or else NAs come up in Date and datetime]
# convert date column to date format
# combine date and time columns 
temp_light_data <- temp_light_data %>%
  mutate(Date = mdy(Date),
  datetime = ymd_hms(paste(Date, as.character(`Time (GMT)`))))

# look at new datetime column in dataset
head(temp_light_data)
# datetime format now = 2017-12-11 15:00:00

str(temp_light_data)
```

3. Using functions from the lubridate package (a lifesaver!), convert that column to datetime format
```{r}
# I used lubridate package functions to combine date and time into a datetime column [ymd_hms] in question 2, I hope that works for this question

# I used the lubridate package within question 2: datetime = ymd_hms(paste(Date, as.character(`Time (GMT)`)))

# I followed along the date and time tutorial on your intro to biology and ecology site
```

4. Make a time series plot of temperature that shows temp (or light) on y and datetime on x. Color by site
```{r}
# time series plot [ x = datetime + y = temp + color by site]
ggplot(temp_light_data, aes(x = datetime, y = temp, color = site)) +
  geom_line() +
  labs(x = "Datetime", y = "Temperature", color = "Site", title = "Temperature Time Series by Site") +
  theme_classic() 
```















